---
title: "Class Notes"
author: "Lancer Brown"
date: "2/28/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Descriptive analysis

**Goal** Describe or summarize a set of data

1. Early analysis when receive new data
2. Generate simple summaries about the samples and their measurements
    + Eg: measures of central tendency or measures of variability
3. NOT for generalizing the results of the analysis to a larger population or trying to make conclusions

## Exploratory analysis

**Goal** Examine the data and find relationships that weren't previously known

1. Explore how different variables might be related
2. Useful for discovering new connections
3. Help to formulate hypotheses and drive the design of future studies and data collection

**Correlation does not imply causation!**

## Inferential analysis

**Goal** Use a relatively small sample of data to say something about the population

1. Provide your estimate of the variable for the population and provide your uncertainty about your estimate
2. Ability to accurately infer information about the larger population depends heavily on sampling scheme

## Predictive analysis

**Goal** Use current and historical data to make predictions about future data

1. Accuracy in predictions is dependent on measure the right variables
2. Many ways to build up prediction models with some being better or worse for specific cases,
    + More data and a simple model generally performs well at predicting future outcomes

**Just because one variable may predict another, it does not mean that one causes the other**

## Causal analysis

**Goal** See what happens to one variable when we manipulate another variable

1. Gold standard in data analysis
2. Often applied to the results of randomized studies that were designed to identify causation
3. Usually analysed in aggregate and observed relationships are usually average effects

## Mechanistic analysis

**Goal** Understand the exact changes in variables that lead to exact changes in other variables

1. Applied to simple situations or those that are nicely modeled by deterministic equations
2. Commonly applied to physical or engineering sciences
    + Eg: Biological sciences, are far too noisy to use mechanistic analysis
3. Often the only noise in the data is measurement error

## Experimental design

Formulate question -> Design -> ID problems and source error -> collect data

Hypothesis: what is outcome of experiment

Y-axis: Dependent variables
X-axis: Independent variables

Example: diet effect on BMI etc. (Diet = dependant)
Confounder: extraneous that could effect relationship dependent vs independent

Reduce confounding effects: controls, blinded, randomization

Replications: repeating experiment with multiple subjects. 

p-value: probability results were observed by chance (p<0.05 - significant result)

p-hacking: exhaustively search for correlations in data in large data sets.

P console Input and Evaluation
x <- 1 assignment operator (x is 1)
print(x)

# indicates comment

x <- ## nothing printed

x ## auto-printing occurs

print(x) ## explicit printing

```{r}
x <- 1
print(x)

msg <- "hello"
msg

## : is used to create integer sequences
x <- 1:20
x

```


## Data Types

5 Basic (atomic) classes of objects
- character
- numeric
- integers
- complex
- logical (True/False)

Most basic object is vector
vector can only contain objects of same class - Cannot Mix 
ONe exception is list: which is a vector but can contain different classes

Empty vectors can be created with vector()

### Numbers

- Numbers are numeric objetcs (double precision real numbers)

- If you explicitly want an integer, you need to specify L suffix.

- Ex. Entering 1 gives you a numeric, entering 1L gives you integer

- There is also a special number Inf which represents infinity: e.g. 1 / 0: Inf can be used in ordinary calculations: e.g. 1 / Inf is 0

- The NaN value represents undefined value "Not a Number" e.g. 0/0; NaN can also be thought of as a missing value (more on that later)

### Attributes

R objects can have attributes

- names, dimnames

- dimensions (e.g. matrices, arrays)

- class

- length

- other user-defined attributes/metadata

Attributes of an object can be accessed using the attributes() function.

## R Data Types - Vectors and Lists

The c() function can be used to create vectors of objects

```{r}
x <- c(0.5, 0.6) ## numeric
x <- c(TRUE, FALSE) ## logical
x <- c(T, F) ## logical
x <- c("a", "b", "c") ## character
x <- 9:29 ## integer
x <- c(1+0i, 2+4i) ## complex

## Using the vector() function
x <- vector("numeric", length = 10)
x
```

## Mixing Objects

What about the following?

```{r}
y <- c(1.7, "a") ## character - first element is 1.7, second element is "a"

y

y <- c(TRUE, 2) ## numeric - TRUE gets converted to a number (TRUE = 1; FALSE = 2); vector 1,2

y

y <- c("a", TRUE) ## character 

```

When different objects are mixed in a vector, coercion occurs so that every element in the vector is of the same class. Coercion goes with least common denominator.

## Explicit Coercion

Objects can be explicitly coerced from one class to another using the as.* functions, if available.

```{r}
x <- 0:6
class(x)

as.numeric(x)

as.logical(x)

as.character(x)

```

Nonsensical coercion results in NAs

```{r}
x <- c("a", "b", "c")
as.numeric(x)

as.logical(x)

as.complex(x)



```


## Lists


Lists are a special type of vector that can contain elements of different classes. Lists are very important data type in R and you should get to know them well.

```{r}
x <- list(1, "a", TRUE, 1+4i)
x

```
Elements are indexed with double brackets; one way to separate list from other types of vectors.

## Matrices

Special type of vector with a *dimension* attribute.The dimension attribute is itself an integer vector of length 2 (nrow, ncol)

```{r}
m <- matrix(nrow = 2, ncol = 3)

dim(m) ## dim is dimension attribute

attributes(m)


```

Matrices are constructed column-wise, so entries can be thought of starting in the "upper left" corner and running down the columns.

```{r}
m <- matrix(1:6, nrow = 2, ncol = 3)
m
```


Matrices can also be created directly from vectors by adding a dimension attribute.

```{r}
m <- 1:10
m

dim(m) <- c(2, 5)
m

```

Matrices can be created by *column-binding* or *row-binding* with *cbind()* or *rbind()*

```{r}
x <- 1:3
y <- 10:12
cbind(x,y)

rbind(x,y)
```

## Factors

Factors are used to represent categorical data. Factors can be unordered or ordered. One can think of a factor as an integer vector where each integer has a *label*

- Factors are treated specially by modelling functions like lm() and glm()
- Using factors with labels is *better* than using integers because factors are self-describing: having a variable that has values "Male" and "Female" is better than a variable that has values 1 and 2. 

Think of a factor as an integer vector where each integer has a label. So for example, you might, you can think of it as a vector as one two three, where one represents you know, high, for example high value and two represents a medium value and three represents a low value. So you might have a, a variable that's called high, medium and low. And underlying in R is represented by the numbers one, two, and three.

Factors can be created by a factor function.

```{r}
x <- factor(c("yes", "yes", "no", "yes", "no"))
x
table(x) # frequence of how many each there are
unclass(x)


```
Factors are integer vectors with levels attribute of "no" and "yes"

The order of the levels can be set using the *levels* argument to factor(). This can be important in linear modeling because the first level is used as the baseline level.

```{r}
x <- factor(c("yes", "yes", "no", "yes", "no"),
            levels = c("yes", "no"))
x

```

Factor order baseline level is determined in R by alphabetical order.


## Missing values

Missing values are denoted by NA or NaN for undefined mathematical operations.

- is.na() is used to test objects if they are NA
- is.nan() is used to test for NaN
- NA values have a class also, so there are integer NA, character NA, etc.
- A NaN value is also NA but the converse is not true.

```{r}
x <- c(1, 2, NA, 10, 3)
is.na(x)
is.nan(x)
x <- c(1, 2, NaN, NA, 4)
is.na(x)
is.nan(x)
```


## Data frames

Data frames are used to store tabular data

- They are represented as a special type of list where every element of the list has to have the same length.

- Each element of the list can be thought of as a column and the length of each element of the list is the number of rows

- Unlike matrices data frames can store different classes of objects in each column (just like lists); matrices must have every element be the same class.

- So the first column could be numbers, the second column could be factor, the third column could be integers the fourth column could be logicals, it doesn't matter what the different types are.

- Data frames also have a special attribute called row.names

- Data frames are usually created by calling read.table() or read.csv()

- Can be convereted to a matrix by calling data.matrix()

```{r}
x <- data.frame(foo = 1:4, bar = c(T,T, F, F))
x
nrow(x)
ncol(x)
```

## Names attribute

R objects can also have names, which is very useful for writing readable code and self-describing objects.

```{r}
x <- (1:3)
names(x)
names(x) <- c("foo", "bar", "norf")
x
names(x)
```

Lists can have names

```{r}
x <- list(a = 1, b = 2,c = 3)
x
```

Matrices can have names

```{r}
m <- matrix(1:4, nrow = 2, ncol = 2)
dimnames(m) <- list(c("a", "b"), c("c", "d"))
m
```

## Summary

Data types

- Atomic classes: numeric, logical, character, integer, complex

- vectors,lists

- factors

- missing values

- data frames

- names

## dput-ting / dget-ting R Objects

Pass data around by deparsing R object with dput and reading it back in using dget.

```{r}
y <- data.frame(a = 1, b = "a")
dput(y)
dput(y, file = "y.R")
new.y <- dget("y.R")
new.y
```


## Removing NA Values

```{r}
x <- c(1, 2, NA, 4, NA, 5)
bad <- is.na(x)
bad
x[!bad]

```

# Matrix Multiplication (true matrix multiplication a %*% b)

1, 3  x 10, 10    = 40, 40  (1x10 + 3x10), (1x10 + 3x10)
2, 4    10, 10      60, 60  (2x10 + 4x10), (2x10 + 4x10)

colMeans(w1ds, na.rm = TRUE, dims = 1)


summarise(y, mean(Solar.R))








