---
title: "Class Notes"
author: "Lancer Brown"
date: "2/28/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Descriptive analysis

**Goal** Describe or summarize a set of data

1. Early analysis when receive new data
2. Generate simple summaries about the samples and their measurements
    + Eg: measures of central tendency or measures of variability
3. NOT for generalizing the results of the analysis to a larger population or trying to make conclusions

## Exploratory analysis

**Goal** Examine the data and find relationships that weren't previously known

1. Explore how different variables might be related
2. Useful for discovering new connections
3. Help to formulate hypotheses and drive the design of future studies and data collection

**Correlation does not imply causation!**

## Inferential analysis

**Goal** Use a relatively small sample of data to say something about the population

1. Provide your estimate of the variable for the population and provide your uncertainty about your estimate
2. Ability to accurately infer information about the larger population depends heavily on sampling scheme

## Predictive analysis

**Goal** Use current and historical data to make predictions about future data

1. Accuracy in predictions is dependent on measure the right variables
2. Many ways to build up prediction models with some being better or worse for specific cases,
    + More data and a simple model generally performs well at predicting future outcomes

**Just because one variable may predict another, it does not mean that one causes the other**

## Causal analysis

**Goal** See what happens to one variable when we manipulate another variable

1. Gold standard in data analysis
2. Often applied to the results of randomized studies that were designed to identify causation
3. Usually analysed in aggregate and observed relationships are usually average effects

## Mechanistic analysis

**Goal** Understand the exact changes in variables that lead to exact changes in other variables

1. Applied to simple situations or those that are nicely modeled by deterministic equations
2. Commonly applied to physical or engineering sciences
    + Eg: Biological sciences, are far too noisy to use mechanistic analysis
3. Often the only noise in the data is measurement error

## Experimental design

Formulate question -> Design -> ID problems and source error -> collect data

Hypothesis: what is outcome of experiment

Y-axis: Dependent variables
X-axis: Independent variables

Example: diet effect on BMI etc. (Diet = dependant)
Confounder: extraneous that could effect relationship dependent vs independent

Reduce confounding effects: controls, blinded, randomization

Replications: repeating experiment with multiple subjects. 

p-value: probability results were observed by chance (p<0.05 - significant result)

p-hacking: exhaustively search for correlations in data in large data sets.

P console Input and Evaluation
x <- 1 assignment operator (x is 1)
print(x)
# indicates comment

```{r}
x <- 1
print(x)

msg <- "hello"
```



